Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2023-08-31 paper/markdown/manuscript.Rmd
Author's feedback
Incorporation of the most recent author's feedback into the manuscript text and Supplementary Material

---
title: "Three distinct patterns of mental response after mountain sport accidents"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author: 
  - Katharina Hüfner, MD:
      institute: [psych]
      email: katharina.huefner@tirol-kliniken.at
      correspondence: true
  - Hanna Salvotti, MD:
      institute: [psych]
  - Piotr Tymoszuk, PhD:
      institute: [daas]

institute:
  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria
  - daas: Data Analysis as a Service Tirol, Innsbruck, Austria

@@

__Background:__ Psychological consequences of mountain sport accidents are poorly characterized. 
We investigated patterns of mental response to mountain sport accidents and their victims at risk of mental deterioration.

__Methods:__ A cross-sectional survey was conducted with `r nrow(ptsd$dataset)` adult victims of mountain sport accidents treated at the Innsbruck University Hospital (Austria). 
Anxiety (GAD-7), depression, somatic symptoms and panic (PHQ), resilient comping (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD) were recorded along with demographic, socioeconomic, medical history, accident- and injury-related data. 
Patterns of mental readouts were defined by semi-supervised medoid clustering and models of cluster assignment based on non-mental measures were developed by the random forest algorithm.

__Results:__ Three mental clusters of accident victims were hallmarked by a neutral mental response, post-traumatic growth and post-traumatic burden of mental disorders, respectively. 
Despite relatively frequent hospitalization, surgery, pre-existing mental or somatic illness, low income, and female gender in the post-traumatic growth and post-traumatic burden clusters, overall differences in non-mental variables between the clusters were weak. 
Consequently, no reliable model of the cluster assignment employing explanatory factors available during acute medical accident management could be established.

__Conclusion:__ A subset of mountain sport accident victims develops symptoms of anxiety, depression, post-traumatic stress disorder and other mental health problems. 
Since identification of these vulnerable patients with demographic, socioeconomic, accident- and injury-related predictors remains challenging, early and low-threshold psychological support is key to prevention of mental disorders following mountain sport accidents.

@@

mountain sport accident, mental health, post-traumatic stress disorder, post-traumatic growth, clustering, random forest, machine learning

@@

@@

The study was conducted in accordance with the Declaration of Helsinki and European data policies. 
All participants gave electronically signed written informed consent to participate. 
Participants' data were processed in anonymized form. 
The study protocol was approved by the ethics committee of the Medical University of Innsbruck (approval number: _please fill in!_).

@@

In- and outpatients treated at the orthopedics and trauma departments of the Medical University of Innsbruck (Austria) between 2018 and 2020 were screened for participation. 
Patients fulfilling the inclusion criteria: age $\geq$ 16 years, proficiency in German and treatment due to a mountain sport accident (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey. 
Out of them, `r length(ptsd$interview_ID)` completed the survey and `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S2__).

@@

The psychometric battery consisted of German versions of assessment tools for 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
resilient coping (RS13: 13-item resilience scale) [@Leppert2008], 
loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__). 
Signs of anxiety (GAD-7 $\geq$ 10), depression (PHQ-9 $\geq$ 11) [@Manea2012], persistent somatic symptoms (PHQ-15 $\geq$ 11) [@Kroenke2002] and resilient coping classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73)[@Leppert2008] were defined as described before. 
The total PTGI and PCL-5 DSM-5 scores were calculated as the sum of all items of the respective tools [@Tedeschi1996; @Bovin2016]. 
Significant PTSD symptoms were defined as positivity in at least one PCL-5 DSM-5 domain [@Bovin2016].

Traumatic events before the accident were assessed with the DIA-X tool [@Maercker2005]. 
Flashbacks frequency was surveyed as none, > 1/month and > 1/year. 
Smoking was surveyed as a single yes/no question. 
Alcohol abuse was investigated with the CAGE tool [@OBrien2008]. 
Data on injury (AIS: abbreviated injury scale) [@Gennarelli2006] and medical treatment were extracted from electronic patient's records.

The training and test participant subsets (2:1 size ratio) were obtained by random splitting which minimized differences in accident year, age, gender, pre-existing traumatic events, somatic and mental illness, and injury severity (__Figure \@ref(fig:fig-consort)__). 
The training subset size (n = `r nrow(clust_globals$analysis_tbl$training)`) was sufficient for clustering analysis as assessed by Hopkins metrics = `r pow$hopkins %>% filter(size == 200) %>% .$hopkins_stat %>% median %>% signif(2)` [@Kassambara2020].

@@

The primary analysis endpoint was identification of subsets of mountain sport accident victims by clustering in respect to scores of anxiety, depression, panic, persistent somatic symptoms, quality of life, post-traumatic growth and post-traumatic stress disorder. 
The secondary analysis endpoints were demographic, socioeconomic and clinical characteristic of the mental clusters and development of a mental cluster assignment model to discern accident victims at risk of mental disorders.

@@

The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Drost2018; @Schubert2019]. 
This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares) and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical and KMEANS algorithms. 
The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic [@Kassambara2020; @Rousseeuw1987]. 
The training subset observations were assigned to the clusters with an inverse distance weighted 7-nearest neighbor classifier.

The uni- and multi-parameter classifiers were trained with the one-rule [@VonJouanne-Diedrich2017; @Holte1993] and conditional random forest algorithms [@Hothorn2006; @Strobl2007; @Hothorn2022], respectively. 
The cluster assignment was predicted for the test subset and assessed by the accuracy and $\kappa$ statistics [@Kuhn2008]. 
Conditional variable importance in the random forest classifiers was expressed as accuracy loss [@Strobl2008].

@@

# Characteristic of the study cohort

In- and outpatient adult victims of mountain sport accidents treated at the Medical University of Innsbruck (Austria) between 2018 and 2020 were invited to participate (n = `r nrow(ptsd$cleared)`). 
Out of them, 
`r length( ptsd$interview_ID)` completed the study survey and 
`r nrow(ptsd$dataset)` individuals with the complete psychometric parameter set were analyzed (overall response rate: `r ((nrow(ptsd$dataset)/nrow(ptsd$cleared)) * 100) %>% signif(2)`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__).

As compared with the participants, the individuals excluded from the analysis had lower income, less frequent ski and more frequent biking accidents, less severe injuries, and lower hospitalization and surgery rates. 
The excluded individuals had also lower ratings of post-traumatic growth ratings and lower sense of coherence. 
Effects of these differences were weak (V $\leq$ 0.22 or r $\leq$ 0.15, __Supplementary Table S4__).

In the analyzed cohort, the median time between the medical accident management and survey completion was 
`r median(ptsd$dataset$obs_time)` days. 
The participants were predominantly middle-aged (median `r median(ptsd$dataset$age)` years) and 
`r signif(table(ptsd$dataset$sex)/sum(table(ptsd$dataset$sex)), 2)[1] * 100`% of them were females. 
The vast majority had secondary or tertiary education grade and were professionally active. 
Sport and trauma-risk professions constituted < 8% of the cohort. 
High annual household incomes of $\geq$ 45000 Euro was reported by > 40% of participants. 
Less than 10% of participants were smokers or at risk of alcohol addiction. 
Chronic somatic conditions prior to the accident were reported by `r signif(table(ptsd$dataset$somatic_comorbidity)/sum(table(ptsd$dataset$somatic_comorbidity)), 2)[2] * 100`% of participants with cardiovascular, neurological and metabolic illness being the most frequent. 
The rates of smoking (Austria: 20.6%) and somatic illness (38.3%) were lower than in the general Austrian population [@StatistikAustria2022a; @StatistikAustria2022].
Psychiatric disorders before the accident affected `r signif(table(ptsd$dataset$psych_comorbidity)/sum(table(ptsd$dataset$psych_comorbidity)), 2)[2] * 100`% of the cohort. 
Four of ten participants were affected by or witnessed a traumatic event prior to the accident (__Table \@ref(tab:tab-cohort-demo)__).

Nearly 40% of participants had sport accidents in the past. 
Almost two-thirds of the investigated accidents happened during skiing or snowboarding followed by biking and classical mountain sports such as climbing, hiking or mountaineering. 
One-third participants were alone during the accident and, in most cases, were the sole culprit and victim. 
Roughly half of participants could rescue themselves, `r signif(table(ptsd$dataset$accident_rescue)/sum(table(ptsd$dataset$accident_rescue)), 2)[3] * 100`% required a professional rescue. 
In `r signif(table(ptsd$dataset$injury_sev_strata)/sum(table(ptsd$dataset$injury_sev_strata)), 2)[2] * 100`% of participants the injury was moderate (AIS 2) and in 
`r signif(table(ptsd$dataset$injury_sev_strata)/sum(table(ptsd$dataset$injury_sev_strata)), 2)[3] * 100`% severe-to-critical (AIS $\geq$ 3). 
Limb injuries were the most common followed by head and face (__Supplementary Figure S1A__). 
Hospitalization and surgery rates were 
`r signif(table(ptsd$dataset$hospitalization)/sum(table(ptsd$dataset$hospitalization)), 2)[2] * 100`% and 
`r signif(table(ptsd$dataset$surgery_done)/sum(table(ptsd$dataset$surgery_done)), 2)[2] * 100`%, respectively. 
Psychological or psychiatric support after the accident was applied to 
`r signif(table(ptsd$dataset$psych_support_post_accident)/sum(table(ptsd$dataset$psych_support_post_accident)), 2)[2] * 100`% individuals. 
Among those who had not received it, 
`r signif(table(ptsd$dataset$psych_support_need)/sum(table(ptsd$dataset$psych_support_need)), 2)[2] * 100`% declared psychological support need. 
Despite fairly common somatic health consequences of the accident 
(`r signif(table(ptsd$dataset$accident_aftermath)/sum(table(ptsd$dataset$accident_aftermath)), 2)[2] * 100`%) and flashbacks during sport (`r signif(table(ptsd$dataset$unwilling_flashback)/sum(table(ptsd$dataset$unwilling_flashback)), 2)[2] * 100`%), most participants returned to the accident sport type (
`r signif(table(ptsd$dataset$same_sport_type_post_accident)/sum(table(ptsd$dataset$same_sport_type_post_accident)), 2)[2] * 100`%). 
Yet, two-third described their behavior during sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).

Signs of 
anxiety (`r signif(table(ptsd$dataset$gad7_total_class)/sum(table(ptsd$dataset$gad7_total_class)), 2)[2] * 100`%), 
depression (`r signif(table(ptsd$dataset$phq9_total_class)/sum(table(ptsd$dataset$phq9_total_class)), 2)[2] * 100`%) and 
persistent somatic symptoms (`r signif(table(ptsd$dataset$phq_events_total_class)/sum(table(ptsd$dataset$phq_events_total_class)), 2)[2] * 100`%) were rare in the study cohort (__Table \@ref(tab:tab-cohort-mental)__). 
Quality of life was rated with median `r median(ptsd$dataset$eurohis_total) %>% signif(2)` EUROHIS points, which was comparable to the generalized Western European population [@Schmidt2006]; overall quality of life and housing domains received the highest ratings. 
Roughly two-thirds of participants were assigned to the high resilient coping class [@Leppert2008]. 
Overall post-traumatic growth (median: `r median(ptsd$dataset$ptgi_total) %>% signif(2)`) and its specific domains were rated lower than originally reported for the PTGI tool [@Tedeschi1996]. 
Symptoms of post-traumatic stress disorder were observed in `r signif(table(ptsd$dataset$dsm5_cluster_class)/sum(table(ptsd$dataset$dsm5_cluster_class)), 2)[2] * 100`% of participants with domain B symptoms (`r signif(table(ptsd$dataset$dsm5_B_class)/sum(table(ptsd$dataset$dsm5_B_class)), 2)[2] * 100`%) being most common (__Table \@ref(tab:tab-cohort-mental)__ and __Supplementary Figure S1__).

For clustering and modeling, the cohort was split into the training (n = `r nrow(clust_globals$analysis_tbl$training)`) and test subset (n = `r nrow(clust_globals$analysis_tbl$test)`). 
There were no significant differences between the subsets in demographic, socioeconomic, clinical and accident-related parameters (not shown).

# Three clusters of mental response in sport accident victims

>>>>

---
title: "Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center"
subtitle: "Manuscript"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

author: 
  - Hanna Salvotti, MD:
      institute: [psych, regensburg]
  - Piotr Tymoszuk, PhD:
      institute: [daas]
  - Mathias Ströhle, MD:
      institute: [kufstein]
  - Peter Paal, MD: 
      institute: [salzburg, kurasi, himed, icar]
  - Hermann Brugger, MD: 
      institute: [eurac, ismm]
  - Martin Faulhaber, PhD: 
      institute: [sport]
  - Nicola Kruijer, MD: 
      institute: [trauma]
  - Barbara Sperner-Unterweger, MD: 
      institute: [psych]
  - Katharina Hüfner, MD:
      institute: [psych]
      email: katharina.huefner@tirol-kliniken.at
      correspondence: true

institute:
  - psych: Department of Psychiatry, Psychotherapy, Psychosomatics and Medical Psychology, University Hospital for Psychiatry II, Medical University of Innsbruck, Innsbruck, Austria.
  - regensburg: Department of Neurosurgery, University Hospital of Regensburg, Regensburg, Germany
  - daas: Data Analysis as a Service Tirol, Innsbruck, Austria
  - kufstein: Department of Anesthesiology and Critical Care Medicine, Bezirkskrankenhaus Kufstein, Kufstein, Austria.
  - salzburg: Department of Anesthesiology and Intensive Care Medicine, Paracelsus Medical University, Salzburg, Austria.
  - kurasi:  Austrian Board of Mountain Safety (Österreichisches Kuratorium fur Alpine Sicherheit), Innsbruck, Austria.
  - himed: Austrian Society of Mountain and High-Altitude Medicine, Mieming, Austria. 
  - icar: International Commission for Mountain Emergency Medicine (ICAR), Kloten, Switzerland.
  - eurac: EURAC Research, Institute for Mountain Emergency Medicine, Bolzano/Bozen, Italy.
  - ismm: International Society of Mountain Medicine (ISMM).
  - sport: Department of Sport Science, University of Innsbruck, Innsbruck, Austria.
  - trauma: Department of Orthopedics and Traumatology, Medical University of Innsbruck, Innsbruck, Austria.

@@

__Background:__ Mountain sports are inherently linked to physical injury risk. 
We aimed to characterize mental health following mountain sport accidents requiring professional medical management.

__Methods:__ Adult victims of mountain sport accidents treated at the Innsbruck University Hospital (Austria) between 2008 and 2020 (n = `r nrow(ptsd$dataset)`) participated in a cross-sectional survey study at median `r round(median(ptsd$dataset$obs_time)/30.436)` months after the trauma center admission. Symptoms of anxiety, depression, somatization and panic disorder (PHQ), resilience (RS-13), sense of coherence (SOC-9L), quality of life (EUROHIS-QOL), post-traumatic growth (PTGI) and post-traumatic stress disorder (PTSD, PCL-5), sociodemographic and clinical information were obtained from an on-line survey and electronic health records. 
Mental health patterns were investigated by semi-supervised medoid clustering and modeled by machine learning.

__Results:__ At least one diagnostic criterion of PTSD was met by `r signif(table(ptsd$dataset$dsm5_cluster_class)[2]/nrow(ptsd$dataset), 2) * 100`% of participants. 
Three comparably sized subsets of patients were identified: (1) neutral, (2) post-traumatic growth and (3) post-traumatic stress cluster. The post-traumatic stress cluster was characterized by frequent mental disorder symptoms, and low resilience, sense of coherence and quality of life as well as by low age, the highest frequency of pre-existing mental disorders and common physical health consequences of the accident. Reliability of machine learning-based prediction of the cluster assignment based on `r length(class_globals$variables$early)` variables available during acute treatment was limited.

__Conclusion:__ A subset of mountain sport accident victims shows symptoms of mental health disorders including PTSD. 
Since early identification of these vulnerable patients remains challenging, low-threshold psychological support is key for a successful interdisciplinary management of mountain sport accidents.

@@

mountain sport accident, mental health, post-traumatic stress disorder, post-traumatic growth, machine learning

@@

Physical activity in mountains, such as hiking, climbing, alpine skiing or ski touring, gain popularity, a trend that has been accelerated by the COVID-19 pandemic [@Hedenborg2022]. 
Outdoor physical activity in general and mountain sport in particular exerts beneficial effects on mental health [@Ower2019; @Niedermeier2017a; @Niedermeier2017]. 
However, mountain sports are also associated with the risk of injury [@Faulhaber2017; @Faulhaber2020], which could foster development of mental disorders including post-traumatic stress disorder (PTSD).

PTSD symptoms include hyperarousal, panic, somatic symptoms, sleep disorders and unwilling recalls of the trauma (flashbacks) [@Bryant2019; @AmericanPsychiatricAssociation2013; @Howie2019], which can occur following self-experiencing or witnessing of a traumatic event such as actual or threatened death, injury, physical or sexual violence. 
PTSD may co-occur with symptoms of depression and anxiety and substance abuse disorders [@Gagnon-Sanschagrin2022; @Brady2000]. 
Although most individuals have experienced at least one traumatic event during the lifetime [@Benjet2016; @Breslau2004], the prevalence of PTSD in the general European population has been estimated for 0.38 to 6.67% [@Trautmann2018; @Burri2014]. 
Yet, up to 26% of acute physical trauma patients have been reported to develop PTSD [@Joseph2020].

So far it has proven difficult to identify early predictors of subsequent PTSD development in physical trauma patients. 
While some recent studies on specific populations have identified risk factors such as pre-existing psychological distress [@Ziobrowski2021] or symptoms of mental disorders [@Papini2018], others have focused on the arousal and stress reaction during acute trauma and peri-traumatic period as candidate predictors [@Schultebraucks2020]. 
These results require validation and further generalization before they can be implemented into routine care [@Wafa2019; @Shalev2019]. 
Many trauma patients at risk of PTSD undergo intensive acute medical management at emergency departments, which poses a challenge for implementation of early diagnostic and psychological preventive measures [@Linares2017]. 
Furthermore, efficacy of such acute psychological interventions at preventing PTSD in trauma patients is uncertain, probably due to large heterogeneity of study cohorts [@Qi2016].

Post-traumatic growth describes a salutary adjustment psychological reaction to a traumatic experience with strengthened perceptions of self, others, and meaning of the event [@Tedeschi1996]. 
Resilience, i.e. the ability to master adversities has been proposed to bridge traumatic events and post-traumatic growth. 
It was also suggested to explain why most people do not develop PTSD despite the high prevalence of traumatic events in the general population [@Ahmed2007; @Bensimon2012]. 
Overall, mountain sport was associated with high resilience values [@Ower2019].

Reports on mental health following accidents during mountain sports are scarce. 
The available data concern mountain guides, rescuers or avalanche accident survivors [@Leonard2021; @Sommer2004; @Mikutta2022; @Thordardottir2015; @Traber2020]. 
Considering the large number of mountain accident victims, and heterogeneity of accident and injuries, accurate characteristic of mental health of the affected patients is urgently needed. 
In particular, tools allowing for identification of patients at risk of later mental health problems during acute medical management of the mountain accident would facilitate the early targeted psychological support and help to preserve the positive mental health effects of physical activity.

Herein, we analyzed a broad palette of mental health readouts such as anxiety, depression, resilience, sense of coherence, quality of life, post-traumatic growth and signs of PTSD in a collective of `r nrow(ptsd$dataset)` patients treated for injuries during mountain sport accidents at the Medical University of Innsbruck, Austria by means of semi-supervised clustering.  
Additionally, we searched for early sociodemographic and clinical predictors of mental health impairment following a mountain accident with a machine learning approach.

@@

The study was conducted in accordance with the Declaration of Helsinki and European data policies. 
All participants gave electronically signed written informed consent to participate. 
Participants' data were processed in anonymized form. 
The study protocol was approved by the ethics committee of the Medical University of Innsbruck (approval number: 1472/2020).

@@

Patients treated for a mountain sport accident at the Department for Orthopedics and Traumatology at the Medical University of Innsbruck between 1^st^ January 2018 and 31^st^ December 2020 (i.e. at least 6 months prior to the start of the study) were screened for participation. 
Patients fulfilling the inclusion criteria: age $\geq$ 18 years, residence in a German-speaking country and proficiency in German (n = `r nrow(ptsd$cleared)`) were invited to participate in the online study survey via conventional mail. 
Out of the invited subjects, `r length(ptsd$interview_ID)` completed the survey and `r nrow(ptsd$dataset)` individuals with the complete psychometry data were analyzed (__Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S2__).

@@

Sociodemographic (e.g. age, sex, education, profession, income, prior sport accidents and traumatic events), accident- and recovery-related (e.g. rescue, psychological support, persistent physical health consequences) and psychometric variables were recorded with a cross-sectional on-line survey.

The psychometric battery consisted of German versions of assessment tools for symptoms of 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ: patient health questionnaire, PHQ-9, depression module) [@Lowe2002; @Grafe2004], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
persistent somatic symptoms (PHQ-15, somatization module) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
lack of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016]. 
The tools displayed good-to-excellent consistency [@McDonald1999] (McDonald's $\omega$ > 0.8, __Supplementary Table S2__ and __S3__).

Clinically significant symptoms of anxiety (GAD-7 $\geq$ 10), 
depression (PHQ-9 $\geq$ 11) [@Manea2012], 
persistent somatic symptoms (PHQ-15 $\geq$ 11) [@Kroenke2002] and 
resilience classes (low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73)[@Leppert2008] 
were defined with published cutoffs.

Separate scores were calculated for domains B, C, D and E of the PCL-5 tool along with the total score being the sum of all items. 
Each PCL-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].

Traumatic events before the accident were assessed with the DIA-X tool [@Maercker2005]. 
Flashbacks frequency was surveyed as none, > 1/year and > 1/month. 
Smoking was surveyed as a single yes/no question. 
Problematic alcohol use was investigated with the CAGE tool [@OBrien2008]. 
Data on the type of the accident mountain sport, injury diagnosis, severity (AIS: abbreviated injury scale) [@Gennarelli2006] and location, hospitalization, surgery and number of surgical ICD-10 diagnoses were extracted from electronic patient's records.

The training and test participant subsets (3:1 size ratio) were obtained by random splitting which minimized differences in in sociodemographic, medical history, clinical and accident- and injury-related variables assessed by Gower distance (__Figure \@ref(fig:fig-consort)__). 
The training subset size (n = `r nrow(clust_globals$analysis_tbl$training)`) was sufficient for clustering analysis as assessed by Hopkins metrics of `r signif(pca$clust_tendency$data$hopkins_stat, 2)` indicative of substantial spontaneous clustering tendency [@Kassambara2020].

@@

The primary analysis endpoint was assessment of mental health of mountain accident victims by clustering using standardized psychometric tools for symptoms of anxiety, depression, panic, persistent somatic symptoms, quality of life, resilience, sense of coherence, post-traumatic growth and PTSD. 
The secondary analysis endpoints were 
(1) to compare demographic, socioeconomic and clinical characteristics between the mental health clusters of mountain accident victims and 
(2) to identify patients at risk of post-traumatic mental disorders based on explanatory variables available during early medical treatment of the mountain sport accident.

@@

The training subset was clustered in respect to normalized median-centered psychometric scores by partition around medoids with cosine distance [@Drost2018; @Schubert2019]. 
This algorithm had a good explanatory performance (ratio of between-cluster sum of squares to total sum of squares) and the superior accuracy in 10-fold cross-validation [@Lange2004] as compared with the hierarchical clustering and KMEANS algorithms. 
The k = 3 cluster number choice was based on the bend of within-cluster sum of squares curve and maximal mean silhouette statistic [@Kassambara2020; @Rousseeuw1987]. 
The training subset observations were assigned to the clusters with an inverse distance-weighted 27-nearest neighbor classifier.

Multi-parameter machine learning classifiers of the mental health cluster assignment were trained with 
the canonical random forest algorithm [@Wright2017; @Breiman2001], 
regularized neural network with a single hidden layer [@Ripley2014], 
support vector machine algorithm with radial kernel [@Karatzoglou2004; @Weston1998], 
recursive partitioning [@Therneau2022; @Breiman2017], 
multinomial elastic net regression [@Zou2005; @Friedman2010], 
and conditional random forest algorithm [@Hothorn2006; @Strobl2007; @Hothorn2022]. 
The cluster assignment was predicted for the test subset and the classifiers' performance at predicting the cluster assignment was assessed by the accuracy and $\kappa$ statistics, Brier score as well as sensitivity and specificity of predicted assignment to the post-traumatic stress cluster [@Cohen1960; @Brier1950; @Kuhn2008].

@@

## Characteristic of the study cohort

The study collective consisted of victims of mountain sport accidents which happened in Tyrol, Austria, a renowned and popular alpine touristic region. 
In- and outpatient adult survivors of mountain sport accidents and treated at the Department of Orthopedics and Traumatology at Medical University of Innsbruck (Tyrol, Austria) between 1^st^ January 2018 and 31^st^ December 2020 were invited to participate (n = `r nrow(ptsd$cleared)`). 
Among the invited patients, 
`r length(ptsd$interview_ID)` completed the study survey and 
`r nrow(ptsd$dataset)` individuals with the complete psychometric data set were analyzed (overall analysis inclusion rate: 
`r ((nrow(ptsd$dataset)/nrow(ptsd$cleared)) * 100) %>% signif(2)`%, __Figure \@ref(fig:fig-consort)__, __Supplementary Tables S1__ - __S3__). 
The median time between the trauma center admission and survey completion was 
`r median(ptsd$dataset$obs_time)` days (interquartile range: `r quantile(ptsd$dataset$obs_time, c(0.25, 0.75)) %>% paste(collapse = ' - ')` days). 
As compared with the analyzed participants, patients who did not respond to the study invitation were characterized by a higher frequency of biking accidents, lower rate of alpine ski/snowboard accidents, less severe injury, lower hospitalization and surgery rates, and less surgery diagnoses (__Supplementary Table S4__). 
The survey participants excluded from the analysis due to missingness of psychometric data had significantly lower income, less severe injuries, more frequent injuries of the upper limbs, less frequently required hospitalization or surgery, and suffered less frequently from persistent physical health consequences of the accident than the analyzed participants (__Supplementary Table S5__). 
Effect sizes of the differences between the included and excluded participants were weak (V $\leq$ 0.22 or r $\leq$ 0.15).

The analyzed participants were predominantly middle-aged 
(median: `r get_num_stats(ptsd$dataset, 'age', collapse = ', interquartile range: ')` years) and 
`r get_percent(ptsd$dataset, 'sex')$percent['female']`% of them were females.
The vast majority had secondary or tertiary education grade 
(`r sum(get_percent(ptsd$dataset, 'education')$percent[c('secondary', 'tertiary')])`%) and were professionally active 
(`r get_percent(ptsd$dataset, 'employment_status')$percent['employed']`%). 
Less than 8% of participants worked in a trauma-exposed or mountain sport profession. 
High annual household incomes of $\geq$ 45000 Euro was reported by 
`r get_percent(ptsd$dataset, 'high_income')$percent['yes']`% of participants. 
Less than 10% of participants were smokers or at risk of problematic alcohol use. 
Pre-existing physical disorders were reported by 
`r get_percent(ptsd$dataset, 'somatic_comorbidity')$percent['yes']`% of participants with cardiovascular, neurological and metabolic illness being the most frequent. 
Four of ten participants had experiences by or witnessed a traumatic event prior to the accident, 
`r sum(get_percent(ptsd$dataset, 'traumatic_number')$percent[c('2', '3+')])`% participants experienced two or more traumatic events. 
Mental disorders prior the accident affected 
`r get_percent(ptsd$dataset, 'psych_comorbidity')$percent['yes']`% of the cohort with affective 
(`r get_percent(mental$analysis_tbl, 'affective')$percent['yes']`%) and somatoform disorders 
(`r get_percent(mental$analysis_tbl, 'somatoform')$percent['yes']`%) as the leading pre-existing mental conditions (__Table \@ref(tab:tab-cohort-demo)__).

Mountain sport accidents in the past were reported by `r get_percent(ptsd$dataset, 'prior_accident')$percent['yes']`% of participants. 
Most of the investigated accidents occurred during alpine skiing, snowboarding or cross-country skiing 
(`r get_percent(ptsd$dataset, 'sport_type')$percent['ski/snowboard/cross-country']`%) followed by biking 
(`r get_percent(ptsd$dataset, 'sport_type')$percent['biking']`%) and classical mountain sports such as climbing, hiking, mountaineering or back-country skiing 
(`r get_percent(ptsd$dataset, 'climbing/hiking/mountaineering/skitour')$percent['climbing/hiking/mountaineering/skitour']`%). 
One-third of participants were alone during the accident 
(`r get_percent(ptsd$dataset, 'accident_alone')$percent['yes']`%) and, in most cases, were the sole culprit 
(`r get_percent(ptsd$dataset, 'accident_culprit')$percent['self']`). 
Professional rescue service was involved in 
`r get_percent(ptsd$dataset, 'accident_rescue')$percent['rescue team']`% of the accidents. 
In `r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['2']`% of participants the injury was moderate (AIS 2) and in 
`r get_percent(ptsd$dataset, 'injury_sev_strata')$percent['3+']`% severe-to-critical (AIS $\geq$ 3). 
Limb injuries were the most common followed by head and face (__Supplementary Figure S1A__). 
Hospitalization and surgery rates following the accident were 
`r get_percent(ptsd$dataset, 'hospitalization')$percent['yes']`% and 
`r get_percent(ptsd$dataset, 'surgery_done')$percent['yes']`%, respectively. 
Psychological or psychiatric support after the accident was provided to 
`r get_percent(ptsd$dataset, 'psych_support_post_accident')$percent['yes']`% individuals. 
A subset of participants (`get_percent(ptsd$dataset, 'psych_support_need')$percent['yes']`%), who had not received it, declared need for psychological support following the accident. 
Despite persisting physical consequences of the accident, which affected 
`r get_percent(ptsd$dataset, 'accident_aftermath')$percent['yes']`% of participants and flashbacks during mountain sport at the time of the questionnaire completion in 
`r get_percent(ptsd$dataset, 'unwilling_flashback')$percent['yes']`% of the cohort, 
most participants returned to the same mountain sport following the accident 
(`r get_percent(ptsd$dataset, 'same_sport_type_post_accident')$percent['yes']`%). 
Yet, `r get_percent(ptsd$dataset, 'caution_post_accident')$percent['more cautious']`% described their behavior during mountain sport as more cautious (__Table \@ref(tab:tab-cohort-accident)__).

At least one diagnostic criterion of PTSD defined by the PCL-5 tool (B, C, D or E) were met by 
`r get_percent(ptsd$dataset, 'dsm5_cluster_class')$percent['positive']`% of participants. 
The domain B PTSD symptoms were the most common (`r get_percent(ptsd$dataset, 'dsm5_B_class')$percent['positive']`%) and the domain D symptoms were the least frequent (`r get_percent(ptsd$dataset, 'dsm5_D_class')$percent['positive']`%). 
Solely `r sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)` patients 
(`r signif(sum(reduce(cohort$ptsd_overlap$data, function(x, y) x * y) > 0)/nrow(cohort$ptsd_overlap$data), 2) * 100`%) were screened positive for all four diagnostic criteria of PTSD (__Table \@ref(tab:tab-cohort-mental)__, __Supplementary Figure S1BC__). 
Clinically relevant symptoms of 
anxiety (`r get_percent(ptsd$dataset, 'gad7_total_class')$percent['positive']`), 
depression (`r get_percent(ptsd$dataset, 'phq9_total_class')$percent['positive']`%) 
and somatic symptoms (`r get_percent(ptsd$dataset, 'phq_events_total_class')$percent['positive']`%) were rare in this cohort consisting of `r get_percent(ptsd$dataset, 'rs13_total_class')$percent['high']`% highly resilient individuals (__Table \@ref(tab:tab-cohort-mental)__).

For clustering and modeling, the cohort was split into the 
training (n = `r table(ptsd$dataset$partition)['training']`) and 
test subset (n = `r table(ptsd$dataset$partition)['test']`). 
Significant differences of weak effect size between these subsets concerned resilience class distribution and scores of quality of life (__Supplementary Table S6__).

## Three clusters of mental response in sport accident victims
